```{r echo=FALSE}
knitr::opts_chunk$set(include=T,echo=T,fig.show=T,results=T,warning=F,message=F,fig.align='center',out.width="75%")
```

# Imagery and Classification Models {#imagery}

```{r echo=F}
library(terra); library(stringr)
imgFolder <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628T230915.",
                    "SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img20mFolder <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R20m")
sentinelBands <- paste0("B",c("02","03","04","05","06","07","8A","11","12"))
imgDateTime <- str_sub(imgFolder,12,26)
sentFiles <- paste0(img20mFolder,"/T10TGK_",imgDateTime,"_",sentinelBands,"_20m.jp2")
sentinel <- rast(sentFiles)
sentFCC <- subset(sentinel,c(7,3,2))
plotRGB(sentFCC,stretch="lin")
```

This chapter explores the use of satellite imagery\index{imagery, satellite}, such as this Sentinel-2\index{Sentinel-2} scene (from 28 June 2021) of part of the northern Sierra Nevada, CA, to Pyramid Lake, NV, including display methods (such as the "false color"\index{false-color} above) and imagery analysis methods. After we've learned how to read the data and cropped it to fit our study area in Red Clover Valley, we'll try some machine language classifiers to see if we can make sense of the patterns of hydric to xeric vegetation in a meadow under active restoration.

Image analysis methods are important for environmental research and are covered much more thoroughly in the discipline of *remote sensing*\index{remote sensing}. Some key things to learn more about are the nature of the electromagnetic spectrum and especially bands of that spectrum\index{electromagnetic spectrum} that are informative for land cover\index{land cover} and especially vegetation\index{vegetation} detection, and the process of imagery classification\index{classification, imagery}. We'll be using satellite imagery\index{satellite imagery} that includes a range of the electromagnetic spectrum from visible to short-wave infrared\index{infrared}.
```{r include=F}
# PLAN:  Include web scraping from Sentinel 2
# maybe using https://cran.r-project.org/web/packages/sen2r/
```

## Reading and Displaying Sentinel-2 Imagery

We'll work with 10 and 20 m Sentinel-2 data downloaded from @Copernicus \index{Copernicus (ESA site for Sentinel imagery)}as an entire scene from 20210602 and 20210628, providing cloud-free imagery of Red Clover Valley \index{Red Clover Valley}in the northern Sierra Nevada\index{Sierra Nevada}, where we are doing research on biogeochemical cycles \index{biogeochemical cycles} and carbon sequestration \index{carbon sequestration}from meadow restoration\index{restoration} with The Sierra Fund (@TSF_CV). 

Download the image zip files from... 

[https://sfsu.box.com/s/zx9rvbxdss03rammy7d72rb3uumfzjbf](https://sfsu.box.com/s/zx9rvbxdss03rammy7d72rb3uumfzjbf)

...and extract into the folder `~/sentinel2/` -- the `~` referring to your Home folder (typically `Documents`). (See the code below to see how one of the two images is read.) Make sure the folders stored there have names ending in `.SAFE`. The following are the two we'll be using:

- `S2B_MSIL2A_20210603T184919_N0300_R113_T10TGK_20210603T213609.SAFE`
- `S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628T230915.SAFE`

```{r}
library(stringr)
library(terra)
imgFolder <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628T230915.",
                    "SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img20mFolder <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R20m")
imgDateTime <- str_sub(imgFolder,12,26)
imgDateTime
imgDate <- str_sub(imgDateTime,1,8)
```


As documented at @Copernicus, Sentinel-2 imagery is collected at three resolutions\index{resolution, imagery}, with the most bands at the coarsest (60 m) resolution. The bands added at that coarsest resolution are not critical for our work as they relate to oceanographic and atmospheric research, and our focus will be on land cover\index{land cover} and vegetation\index{vegetation} in a terrestrial area. So we'll work with four bands\index{bands, imagery} at 10 m and an additional six bands at 20 m resolution:

10 m bands

-   B02 - Blue 0.490 $\mu$m
-   B03 - Green 0.560 $\mu$m
-   B04 - Red 0.665 $\mu$m
-   B08 - NIR 0.842 $\mu$m

20 m bands

-   B02 - Blue 0.490 $\mu$m
-   B03 - Green 0.560 $\mu$m
-   B04 - Red 0.665 $\mu$m
-   B05 - Red Edge 0.705 $\mu$m
-   B06 - Red Edge 0.740 $\mu$m
-   B07 - Red Edge 0.783 $\mu$m
-   B8A - NIR 0.865 $\mu$m
-   B11 - SWIR 1.610 $\mu$m
-   B12 - SWIR 2.190 $\mu$m

```{r}
sentinelBands <- paste0("B",c("02","03","04","05","06","07","8A","11","12"))
sentFiles <- paste0(img20mFolder,"/T10TGK_",imgDateTime,"_",
                        sentinelBands,"_20m.jp2")
sentinel <- rast(sentFiles)
sentinelBands
names(sentinel) <- sentinelBands
```


### Individual bands

\index{imagery band display}The Sentinel-2 data is in a 12-bit format, ranging as integers from 0 to 4,095, but are preprocessed to represent reflectances at the base of the atmosphere.  

We'll start by looking at four key bands -- green, red, NIR, SWIR (B11) -- and plot these in a gray scale (Figure \@ref(fig:img4bands)). 

```{r img4bands, fig.cap=paste0("Four bands of a Sentinel-2 scene from ",imgDate,".")}
par(mfrow=c(2,2))
plot(sentinel$B03, main = "Green", col=gray(0:4095 / 4095))
plot(sentinel$B04, main = "Red", col=gray(0:4095 / 4095))
plot(sentinel$B8A, main = "NIR", col=gray(0:4095 / 4095))
plot(sentinel$B11, main = "SWIR (B11)", col=gray(0:4095 / 4095))
```


### Spectral subsets to create three-band R-G-B and NIR-R-G for visualization
\index{spectral subsetting of imagery for visualization}Displaying the imagery three bands at a time (displayed as RGB\index{RGB} on our computer screen) is always a good place to start (Figure \@ref(fig:imgRGBscene)), and two especially useful band sets are RGB itself -- so looking like a conventional color aerial photography -- and "false color" that includes a band normally invisible to our eyes, such as near infrared that reflects chlorophyll in healthy plants. In standard "false color"\index{false color}, surface-reflected near-infrared is displayed as red, reflected red is displayed as green, and green is displayed as blue\index{NIR-R-G as RGB}. Some advantages of this false color display include:

- Blue is not included at all, which helps reduce haze. 
- Water bodies absorb NIR, so appear very dark in the image. 
- Chlorophyll\index{chlorophyll} strongly reflects NIR, so healthy vegetation is bright red.

```{r imgRGBscene, fig.cap=paste0("R-G-B image from Sentinel-2 scene ",imgDate,".")}
sentRGB <- subset(sentinel,3:1)
plotRGB(sentRGB,stretch="lin")
```
We can also spectrally subset to get NIR (B8A, the 7th band of the 20 m multispectral raster we've created). See the first figure in this chapter for the NIR-R-G image displayed with plotRGB.
```{r fig.cap=paste0("NIR-R-G image from Sentinel-2 scene ",imgDate,".")}
sentFCC <- subset(sentinel,c(7,3,2))
```

### Crop to study area extent
\index{cropping imagery}The Sentinel-2 scene covers a very large area, more than 100 km on each side, 12,000 km^2^ in area. The area we are interested in -- Red Clover Valley -- is enclosed in an area of about 100 km^2^, so we'll crop the scene to fit this area (Figures \@ref(fig:imgRCVcropRGB) and \@ref(fig:imgFalseRCV)).

```{r imgRCVcropRGB, fig.cap=paste0("Color image from Sentinel-2 of Red Clover Valley, ",imgDate,".")}
RCVext <- ext(715680,725040,4419120,4429980)
sentRCV <- crop(sentinel,RCVext)
sentRGB <- subset(sentRCV,3:1)
plotRGB(sentRGB,stretch="lin")
```


**NIR-R-G:**
\index{NIR-R-G as RGB}
```{r imgFalseRCV, fig.cap=paste0("NIR-R-G image from Sentinel-2 of Red Clover Valley, ",imgDate,".")}
plotRGB(subset(sentRCV,c(7,3,2)),stretch="lin")
```


### Saving results
\index{writing rasters}\index{writeRaster, terra}Optionally, you may want to save this cropped image to call it up quicker at a later time. We'll build a simple name with the date and time as part of it.
```{r eval=F}
writeRaster(sentRCV, filename=paste0("~/sentinel2/",imgDateTime,".tif"),overwrite=T)
```

### Band scatter plots
\index{imagery band scatter plots}Using scatter plots helps us to to understand how we might employ multiple bands in our analysis. Red vs NIR is particularly intriguing, commonly creating a triangular pattern reflecting the high absorption of NIR by water yet high reflection by healthy vegetation which is visually green, so low in red (Figure \@ref(fig:imgRedNIRpairs)).
```{r imgRedNIRpairs, fig.cap=paste0("Relations between Red and NIR bands, Red Clover Valley Sentinel-2 image, ",imgDate)}
pairs(c(sentRCV$B04,sentRCV$B8A), main="Red vs NIR")
```

## Spectral Profiles
\index{spectral profiles}A useful way to look at how our imagery responds to known vegetation/land-cover at *imagery training polygons* we'll use later for supervised classification, we can extract values at points and create a *spectral profile* (Figure \@ref(fig:imgSpectralSignatures9)). We start by deriving points from the polygons we created based on field observations. One approach is to use polygon centroids (with `centroids()`), but to get more points from larger polygons, we'll instead sample the polygons to get a total of ~1,000 points; but we'll also set a Boolean variable `doCentroids` to possibly choose this option for this and all other samples in this notebook, if desired.
```{r message=F, warning=F}
library(igisci); library(tidyverse)
trainPolys9 <- vect(ex("RCVimagery/train_polys9.shp"))
doCentroids <- F
if (doCentroids) {ptsRCV <- centroids(trainPolys9)
            }else{ptsRCV <- spatSample(trainPolys9, 1000, method="regular")}
extRCV <- terra::extract(sentRCV, ptsRCV)
head(extRCV)
```

```{r imgSpectral9}
specRCV <- aggregate(extRCV[,-1], list(ptsRCV$class), mean) |>
  rename(cover=Group.1)
LCcolors <- RColorBrewer::brewer.pal(length(unique(trainPolys9$class)),"Set1")
```
```{r}
pivot_specRCVdf <- specRCV |>
  pivot_longer(cols=B02:B12, names_to="band", values_to="reflectance")
```
```{r imgSpectralSignatures9,  fig.cap=paste0("Spectral signature of nine-level training polygons, 20 m Sentinel-2 imagery from ",imgDate,".")}
ggplot(data=pivot_specRCVdf, aes(x=band,y=reflectance,col=cover,group=cover)) + geom_line(size=1.5) +
  labs(y="Reflectance 12 bit") +
  scale_color_discrete(type=LCcolors)
```


One thing we can observe in the spectral signature \index{spectral signature}is that some vegetation types are pretty similar; for instance `mesic`\index{mesic} and `mesicRush` are very similar, as are `hydric`\index{hydric} and `CARNEB`, so we might want to combine them, and have fewer categories.  We also have a seven-level polygon classification, so we can build a reasonable color scheme (for the figure above, we just let the color system come up with as many colors as classes (9)) (Figure \@ref(fig:imgSpectralSignatures7)). 

```{r imgSpectralSignatures7, fig.cap=paste0("Spectral signature of six-level training polygons, 20 m Sentinel-2 imagery from ",imgDate)}
trainPolys7 <- vect(ex("RCVimagery/train_polys7.shp"))
if (doCentroids) {ptsRCV <- centroids(trainPolys7)
            }else{ptsRCV <- spatSample(trainPolys7, 1000, method="regular")}
extRCV <- terra::extract(sentRCV, ptsRCV)
specRCV <- aggregate(extRCV[,-1], list(ptsRCV$class), mean)
specRCVdf <- specRCV|> rename(cover=Group.1)
coverColors <- c("darkgreen","cyan", "orange","gray","blue","green","red")
pivot_specRCVdf <- specRCVdf |>
  pivot_longer(cols=B02:B12, names_to="band", values_to="reflectance")
ggplot(data=pivot_specRCVdf, aes(x=band,y=reflectance,col=cover,group=cover)) + geom_line(size=1.5) +
  labs(y="Reflectance 12 bit") +
  scale_color_discrete(type=coverColors)
  
```

From this spectral signature, we can see that at least at this scale, there's really no difference between the spectral response `hydric` and `willow`. The structure of the willow copses can't be distinguished from the herbaceous hydric sedges with 20 m pixels. Fortunately, in terms of *hydrologic* response, willows\index{willows} are also hydric\index{hydric plants} plants so we can just recode `willow` as `hydric` (Figure \@ref(fig:imgSpectralSignatures6)).

```{r imgtrainingExtract}
library(tidyverse)
trainPolys6 <- trainPolys7
trainPolys6[trainPolys6$class=="willow"]$class <- "hydric"
if (doCentroids) {ptsRCV <- centroids(trainPolys6)
            }else{ptsRCV <- spatSample(trainPolys6, 1000, method="regular")}
extRCV <- terra::extract(sentRCV, ptsRCV)
specRCV <- aggregate(extRCV[,-1], list(ptsRCV$class), mean)
specRCVdf <- specRCV|> rename(cover=Group.1)
```

```{r imgSpectralSignatures6, fig.cap=paste0("Spectral signature of six-level training polygons, 20 m Sentinel-2 imagery from ",imgDate)}
coverColors <- c("darkgreen","green","orange","gray","blue","red")
pivot_specRCVdf <- specRCVdf |>
  pivot_longer(cols=B02:B12, names_to="band", values_to="reflectance")
ggplot(data=pivot_specRCVdf, aes(x=band,y=reflectance,col=cover,group=cover)) + geom_line(size=1.5) +
  labs(y="Reflectance 12 bit") +
  scale_color_discrete(type=coverColors)
  
```

## Map Algebra and Vegetation Indices

\index{map algebra, vegetation indices}We can also do pixel-wise calculations, the *map algebra* method (@Tomlin) used in raster GIS.  We'll use it to create vegetation indices, which are widely used as a single quantitative measure to relate to phenological change, moisture, and nutrients.

### Vegetation indices

\index{vegetation indices}\index{Normalized Difference Vegetation Index NDVI}The venerable *Normalized Difference Vegetation Index* (NDVI) uses NIR (Sentinel2 20m: B8A) and RED (Sentinel2 20m: B04) in a ratio that normalizes NIR with respect to visible (usually red) (Figure \@ref(fig:imgNDVI)).

$$
NDVI = \frac{NIR-RED}{NIR+RED}
$$

```{r imgNDVI, fig.cap=paste0("NDVI from Sentinel-2 image, ",imgDate)}
ndviRCV <- (sentRCV$B8A - sentRCV$B04)/
           (sentRCV$B8A + sentRCV$B04)
plot(ndviRCV, col=rev(terrain.colors(10)), main=paste('NDVI',imgDateTime))
```

### Histogram

\index{histogram (NDVI)}The histogram for an NDVI is ideally above zero as this one is (Figure \@ref(fig:imgNDVIhisto)), with the values approaching 1 representing pixels with a lot of reflection in the near infrared, so typically healthy vegetation. Standing water can cause values to go negative, but with 20 m pixels there's not enough water in the creeks to create a complete pixel of open water. 

```{r imgNDVIhisto, fig.cap=paste0("NDVI histogram, Sentinel-2 image, ",imgDate)}
hist(ndviRCV, main='Normalized Difference Vegetation Index', xlab='NDVI',
     ylab='Frequency',col='#6B702B',xlim=c(-1,1),breaks=40,xaxt='n')
axis(side=1,at=seq(-1,1,0.2), labels=seq(-1,1,0.2))
```

### Other vegetation indices

With the variety of spectral bands available on satellite imagery comes a lot of possibilities for other vegetation (and non-vegetation) indices.  

\index{Normalized Difference Moisture Index NDMI}One that has been used to detect moisture content in plants is the Normalized Difference Moisture Index $NDMI = \frac{NIR-SWIR}{NIR+SWIR}$, which for Sentinel-2 20 m imagery would be $\frac{B8A-B11}{B8A+B11}$ (Figure \@ref(fig:imgNDMIrcv20)).

```{r imgNDMIrcv20, fig.cap=paste0("NDMI from Sentinel-2 image, ",imgDate)}
ndmiRCV <- (sentRCV$B8A - sentRCV$B11)/
           (sentRCV$B8A + sentRCV$B11)
plot(ndmiRCV, col=rev(terrain.colors(10)), main=paste('NDMI',imgDateTime))
```

In contrast to NDVI, the NDMI histogram extends to negative values (Figure \@ref(fig:imgNDMIrcv20histo)). 
```{r imgNDMIrcv20histo, fig.cap=paste0("NDMI histogram, Sentinel-2 image, ",imgDate)}
hist(ndmiRCV, main='Normalized Difference Moisture Index', xlab='NDMI',
     ylab='Frequency',col='#26547C',xlim=c(-1,1),breaks=40,xaxt='n')
axis(side=1,at=seq(-1,1,0.2), labels=seq(-1,1,0.2))
```

Another useful index -- Normalized Difference Greenness Index (NDGI) \index{normalized difference greenness index NDGI}uses three bands: NIR, Red, and Green (@Yang2019), and was first proposed for MODIS but can be varied slightly for Sentinel-2. It has the advantage of using the more widely available four-band imagery, such as is available in the 10 m Sentinel-2 product, 3 m PlanetScope, or multispectral drone imagery using a MicaSense, which for their Altum camera at 120 m above ground is 5.4 cm.  We'll stick with the 20 m Sentinel-2 product for now and produce a plot (Figure \@ref(fig:imgNDGIrcv20)) and a histogram (Figure \@ref(fig:imgNDGIrcv20histo)).

$$
NDGI=\frac{0.616GREEN+0.384NIR-RED}{0.616GREEN+0.384NIR+RED}
$$
```{r}
ndgiRCV <- (0.616 * sentRCV$B03 + 0.384 * sentRCV$B8A - sentRCV$B04)/
           (0.616 * sentRCV$B03 + 0.384 * sentRCV$B8A + sentRCV$B04)
```

```{r imgNDGIrcv20, fig.cap=paste0("NDGI from Sentinel-2 image, ",imgDate)}
plot(ndgiRCV, col=rev(terrain.colors(10)), main=paste('NDGI',imgDateTime))
```

```{r imgNDGIrcv20histo, fig.cap=paste0("NDGI histogram, Sentinel-2 image, ",imgDate)}
hist(ndgiRCV, main='Normalized Difference Greenness Index', 
     xlab='NDGI',ylab='Frequency',
     col='#38635A',xlim=c(-1,1),breaks=40,xaxt='n')
axis(side=1,at=seq(-1,1,0.2), labels=seq(-1,1,0.2))
```

## Unsupervised Classification with k-means

\index{unsupervised k-means classification}We'll look at supervised classification next, where we'll provide the classifier known vegetation or land cover types, but with unsupervised we just provide the number of classes.  We'll use the NDGI we just derived above and start by creating a vector from the values:

```{r}
vals <- values(ndgiRCV)
str(vals)
```

We'll then run the kmeans function to derive clusters:

```{r}
kmc <- kmeans(na.omit(vals), centers=5, iter.max=500, nstart=5, algorithm="Lloyd")
```

Now we convert the vector \@ref(vector) back into a matrix \@ref(matrix) with the same dimensions as the original image...

```{r}
kimg <- ndgiRCV
values(kimg) <- kmc$cluster
```

... and map the unsupervised classification (Figure \@ref(fig:imgUnsupervised20m)). For the five classes generated, we'd need to follow up by overlaying with known land covers to see what it's picking up, but we'll leave that for the supervised classification next.

```{r imgUnsupervised20m, fig.cap=paste0("Unsupervised k-means classification, Red Clover Valley, Sentinel-2, ",imgDate)}
library(RColorBrewer)
LCcolors <- brewer.pal(length(unique(values(kimg))),"Set1")
plot(kimg, main = 'Unsupervised classification', col = LCcolors, type="classes")
```

## Machine Learning Classification of Imagery

\index{machine learning}\index{supervised classification of imagery}Using machine learning algorithms is one approach to classification, whether that classification is of observations in general or pixels of imagery. We're going to focus on a *supervised classification* method to identify land cover from imagery or other continuous raster variables (such as elevation or elevation-derived rasters such as slope, curvature, and roughness), employing samples of those variables called training samples.

It's useful to realize that the modeling methods we use for this type of imagery classification are really no different at the core from methods we'd use to work with continuous variables that might not even be spatial. For example, if we leave off the classification process for a moment, a machine learning algorithm might be used to predict a response result from a series of predictor variables, like predicting temperature from elevation and latitude, or acceleration might be predicted by some force acting on a body. So the first model might be used to predict the temperature of any location (within the boundary of the study area) given an elevation and latitude; or the second model might predict an acceleration given the magnitude of a force applied to the body (maybe of a given mass). A classification model varies on this by predicting a nominal variable like type of land cover; some other types of responses might be counts (using a Poisson model) or probabilities (using a logistic model).

The imagery classification approach adds to this model an input preparation process and an output prediction process:

-   A training set of points and polygons are created that represent areas of known classification such as land cover like forest or wetland, used to identify values of the predictor variables (e.g. imagery bands).
-   A predicted raster is created using the model applied to the original rasters.

\index{classification and regression trees CART}We'll use a machine learning model called **Classification and Regression Trees** (CART) that is one of the simplest to understand in terms of the results it produces, which includes a display of the resulting classification decision "tree".  We have collected field observations of vegetation/land cover types in our meadow, so can see if we can use these to train a classification model. You'll recognize our vegetation types (Figure \@ref(fig:imgTrainingPolys) and \@ref(fig:imgTrainingFill)) from the spectral signatures above. (We'll start by reading in the data again so this section doesn't depend on the above code.) To best see the locations of training samples, many of which are small compared to the overall map extent, two maps are used: first with an imagery basemap, then color-filled over a simple basemap.

### Read imagery and training data and extract sample values for training
```{r}
library(terra); library(stringr); library(tmap)
imgFolder <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628T230915.",
                    "SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img20mFolder <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R20m")
imgDateTime <- str_sub(imgFolder,12,26)
sentinelBands <- paste0("B",c("02","03","04","05","06","07","8A","11","12"))
sentFiles <- paste0(img20mFolder,"/T10TGK_",imgDateTime,"_",sentinelBands,
                    "_20m.jp2")
sentinel <- rast(sentFiles)
names(sentinel) <- sentinelBands
RCVext <- ext(715680,725040,4419120,4429980)
sentRCV <- crop(sentinel,RCVext)
trainPolys <- vect(ex("RCVimagery/train_polys7.shp"))
trainPolys[trainPolys$class=="willow"]$class <- "hydric"
if (doCentroids) {ptsRCV <- centroids(trainPolys)
            }else{ptsRCV <- spatSample(trainPolys, 1000, method="regular")}

```
```{r}
library(maptiles)
trainPolysSF <- sf::st_as_sf(trainPolys)
basemap <- get_tiles(trainPolysSF, provider="Esri.WorldImagery")
bounds=tmaptools::bb(trainPolysSF, ext=1.05)
trainImg <- tm_graticules(alpha=0.5, col="gray") +
  tm_shape(basemap,bbox=bounds) + tm_rgb() + # tm_mv("red","green","blue")) +  
  # tm_basemap("Esri.WorldImagery", alpha=0.5) + # Esri.WorldGrayCanvas Esri.WorldImagery
  tm_shape(trainPolysSF) + 
    tm_polygons(fill=NA, col="class",
                col.scale=tm_scale_categorical(values=coverColors),
                col.legend=tm_legend(frame=F))
```

```{r}
train <- tm_graticules(alpha=0.5, col="gray") +
  # tm_basemap("Esri.WorldGrayCanvas", alpha=0.5) + # Esri.WorldGrayCanvas Esri.WorldImagery
  tm_shape(trainPolysSF) + 
    tm_polygons(fill="class", col="class", #title=NA
                fill.scale=tm_scale_categorical(values=coverColors),
                fill.legend=tm_legend(frame=F),
                col.legend=tm_legend(show=F),
                col.scale=tm_scale_categorical(values=coverColors))
```
```{r imgTrainingPolys, echo=F, fig.cap="Training samples over imagery basemap"}
trainImg
```
```{r imgTrainingFill, echo=F, fig.cap="Training samples over simple basemap"}
train
```

```{r}
LCclass <- c("forest","hydric","mesic","rocky","water","xeric")
classdf <- data.frame(value=1:length(LCclass),names=LCclass)
extRCV <- terra::extract(sentRCV, ptsRCV)[,-1] # REMOVE ID
sampdataRCV <- data.frame(class = ptsRCV$class, extRCV)
```

### Training the CART model
\index{rpart}The *Recursive Partitioning and Regression Trees* package `rpart` fits the model. 
```{r}
library(rpart)
cartmodel <- rpart(as.factor(class)~., data = sampdataRCV, 
                   method = 'class', minsplit = 5)
```

Now we can display the decision tree using the `rpart.plot` (Figure \@ref(fig:imgCARTtree20m)). This function clearly shows where each variable (in our case imagery spectral bands) contributes to the classification, and together with the spectral signature analysis above, helps us evaluate how the process works and possibly how to improve our training data where distinction among classes is unclear.

```{r imgCARTtree20m, fig.cap=paste0("CART Decision Tree, Sentinel-2 20 m, date ",imgDate)}
library(rpart.plot)
rpart.plot(cartmodel, fallen.leaves=F)
```

### Prediction using the CART model

\index{CART prediction}Then we can use the model to predict the probabilities of each vegetation class for each pixel (Figure \@ref(fig:imgCARTclassProb20m)).

```{r imgCARTclassProb20m, fig.cap=paste0("CART classification, probabilities of each class, Sentinel-2 20 m ",imgDate)}
CARTpred <- predict(sentRCV, cartmodel, na.rm = TRUE)
CARTpred
plot(CARTpred)
```

*Note the ranges of values, each extending to values approaching 1.0.* If we had used the nine-category training set described earlier, we would find some categories with very low scores, such that they will never dominate a pixel. We can be sure from the above that we'll get all categories used in the final output, and this will avoid potential problems in validation.

Now we'll make a single SpatRaster showing the vegetation/land cover with the *highest probability* (Figure \@ref(fig:imgCARThighestProb20m)).

```{r}
LC20m <- which.max(CARTpred)
LC20m
cls <- names(CARTpred)
df <- data.frame(id = 1:length(cls), class=cls)
levels(LC20m) <- df
LC20m
```
```{r eval=F, include=F}
LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
plot(LC20m, col=LCcolors)
```

```{r imgCARThighestProb20m, fig.height=6, fig.width=6, fig.cap=paste0("CART classification, highest probability class, Sentinel-2 20 m ",imgDate)}
LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
  # tm_basemap("Esri.WorldGrayCanvas", alpha=0.5) + # Esri.WorldGrayCanvas Esri.WorldImagery
  tm_shape(LC20m) + 
    tm_raster(col="class",  
              col.scale=tm_scale_categorical(values=LCcolors),
              col.legend=tm_legend(frame=F)) +
    tm_graticules(alpha=0.5, col="gray")

```

We could look at landscape metrics, like the Shannon Diversity Index. 

```{r}
library(landscapemetrics)
# landscape <- terra::rast(landscapemetrics::landscape)
lsm_l_shdi(LC20m)
```
```{r}
lsm_c_pland(LC20m)
```


To save the raster, we can write out the classification raster. I haven't figured out how to get the attribute table written yet, but you at least get integers this way.
```{r eval=F}
# raster::ratify(LC20m, count=T)
# foreign::write.dbf(levels(LC20m)[[1]], file="data/LC20m.tif.vat.dbf")
terra::writeRaster(LC20m, "data/LC20m.tif", datatype="INT1U", gdal=c("COMPRESS=NONE", "TFW=YES"), overwrite=T)

```



### Validating the model

\index{validation, model}An important part of the imagery classification process is *validation*, where we look at how well the model works. The way this is done is pretty easy to understand, and requires having *testing* data in addition to the *training* data mentioned above. Testing data can be created in a variety of ways, commonly through field observations but also with finer resolution imagery like drone imagery. Since this is also how we *train* our data, often you're selecting some for training, and a separate set for testing.

#### The "overfit model" problem

\index{overfit model problem}It's important to realize that the accuracy we determine is *only* based on the training and testing data. The accuracy of the prediction of the classification elsewhere will likely be somewhat less than this, and if this is substantial our model is "overfit". We don't actually know how overfit a model truly is because that depends on how likely the conditions seen in our training and testing data also occur throughout the rest of the image; if those conditions are common, just not sampled, then the model might actually be pretty well fit.

In thinking about the concept of overfit models and selecting out training (and testing) sets, it's useful to consider the purpose of our classification and how important it is that our predictions are absolutely reliable. In choosing training sets, accuracy is also important, so we will want to make sure that they are good representatives of the land cover type (assuming that's our response variable). While some land covers are pretty clear (like streets or buildings), there's a lot of fuzziness in the world: you might be trying to identify wetland conditions based on the type of vegetation growing, but in a meadow you can commonly find wetland species mixed in with more mesic species -- to pick a reliable wetland sample we might want to only pick areas with only wetland species (and this can get tricky since there are many challenges of "obligate wetland" or "facultative wetland" species.) The CART model applied a probability model for each response value, which we then used to derive a single prediction based on the maximum probability.

#### Cross validation : 20 m Sentinel-2 CART model

\index{cross validation}In *cross-validation*, we use one set of data and run the model multiple times, and for each validating with a part of the data not used for training the model.  In *k-fold cross validation*, k represents the number of groups and number of models. The k value can be up to the number of observations, but you do need to consider processing time, and you may not get much more reliable assessments with using all of the observations.  We'll use five folds.

The method we'll use to cross-validate the model and derive various accuracy metrics is based on code provided in the "Remote Sensing with *terra*" section of https://rspatial.org by Aniruddha Ghosh and Robert J. Hijmans (@rspatial).

```{r}
set.seed(42)
k <- 5 # number of folds
j <- sample(rep(1:k, each = round(nrow(sampdataRCV))/k))
table(j)
```

```{r}
x <- list()
for (k in 1:5) {
    train <- sampdataRCV[j!= k, ]
    test <- sampdataRCV[j == k, ]
    cart <- rpart(as.factor(class)~., data=train, method = 'class',
                  minsplit = 5)
    pclass <- predict(cart, test, na.rm = TRUE)
    # assign class to maximum probability
    pclass <- apply(pclass, 1, which.max)
    # create a data.frame using the reference and prediction
    x[[k]] <- cbind(test$class, as.integer(pclass))
}
```

#### Accuracy and error metrics
Now that we have our test data with observed and predicted values, we can derive accuracy metrics. There have been many metrics developed to assess model classification accuracy and error. We can look at accuracy either in terms of how well a model works or what kinds of errors it has; these are complements. We'll look at *Producer's* and *User's* accuracies (and their complements *omission* and *commission* errors) below, and we'll start by building a *confusion matrix*, which will allow us to derived the various metrics.

**Confusion Matrix**
\index{confusion matrix}One way to look at (and, in the process, *derive*) these metrics is to build a *confusion matrix*, which puts the **observed** *reference data* and **predicted** *classified data* in rows and columns. These can be set up in either way, but in both cases the diagonal shows those correctly classified, with the row and column names from the same list of land cover classes that we've provided to train the model.

```{r}
y <- do.call(rbind, x)
y <- data.frame(y)
colnames(y) <- c('observed', 'predicted')
conmat <- table(y) # confusion matrix
#str(conmat)
```
```{r}
# change the name of the columns and rows to LCclass
if(length(LCclass)==dim(conmat)[2]){
  colnames(conmat) <- LCclass}else{
    print("Column dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
if(length(LCclass)==dim(conmat)[1]){
  rownames(conmat) <- LCclass}else{
    print("Dimension/array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
```
```{r}
print(conmat)
```

**Overall Accuracy**
\index{overall accuracy}The overall accuracy is simply the overall ratio of the number correctly classified divided by the total, so the sum of the diagonal divided by n.  
```{r}
n <- sum(conmat) # number of total cases
diag <- diag(conmat) # number of correctly classified cases per class
OA20 <- sum(diag) / n # overall accuracy
OA20
```

**kappa**
\index{kappa statistic}Another overall measure of accuracy is *Cohen's kappa* statistic (@CohenKappa), which measures *inter-rater reliability* for categorical data. It's somewhat similar to the goal of chi square in that it compares *expected* vs. *observed* conditions, only chi square is limited to two cases or Boolean data.   
```{r}
rowsums <- apply(conmat, 1, sum) 
p <- rowsums / n # observed (true) cases per class
colsums <- apply(conmat, 2, sum)
q <- colsums / n # predicted cases per class
expAccuracy <- sum(p*q)
kappa20 <- (OA20 - expAccuracy) / (1 - expAccuracy)
kappa20
```

**Table of Producer's and User's accuracies**

\index{Producers and Users accuracies}We can look at accuracy either in terms of how well a model works or what kinds of errors it has; these are complements. Both of these allow us to look at accuracy of classification *for each class*. 

- The so-called "Producer's Accuracy" refers to how often real conditions on the ground are displayed in the classification, from the perspective of the map producer, and is the complement of errors of *omission*\index{omission errors}.
- The so-called "User's Accuracy" refers to how often the class will actually occur on the ground, so it is from the perspective of the user (though this is a bit confusing) and is the complement of errors of *commission*\index{commission errors}.  
 
```{r}
PA <- diag / colsums # Producer accuracy
UA <- diag / rowsums # User accuracy
if(length(PA)==length(UA)){
  outAcc <- data.frame(producerAccuracy = PA, userAccuracy = UA)
  print(outAcc)}else{
     print("length(PA)!=length(UA)")
     print(PA)
     print(UA)
  }
```

## Classifying with 10 m Sentinel-2 Imagery

\index{Sentinel-2 10 m classification}The 10 m product of Sentinel-2 has much fewer bands (4), but the finer resolution improves its visual precision. The four bands are the most common baseline set used in multispectral imagery: blue (B02, 490 nm), green (B03, 560 nm), red (B04, 665 nm), and NIR (B08, 842 nm). From this we can create color and NIR-R-G images, and the vegetation indices NDVI and NDGI, among others.

For the 10 m product, we won't repeat many of the exploration steps we looked at above, such as the histograms, but go right to training and classification using CART. 

```{r}
imgFolder <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628T230915.",
                    "SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img10mFolder <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R10m")
imgDateTime <- str_sub(imgFolder,12,26)
imgDateTime
imgDate <- str_sub(imgDateTime,1,8)
```

### Subset bands (10 m)
The 10-m Sentinel-2 data includes only four bands blue (B02), green (B03), red (B04), and NIR (B08).
```{r}
sentinelBands <- paste0("B",c("02","03","04","08"))
sentFiles <- paste0(img10mFolder,"/T10TGK_",imgDateTime,"_",sentinelBands,
                    "_10m.jp2")
sentinel <- rast(sentFiles)
names(sentinel) <- sentinelBands
```

### Crop to RCV extent and extract pixel values 
This is similar to what we did for the 20 m data.
```{r}
RCVext <- ext(715680,725040,4419120,4429980)
sentRCV <- crop(sentinel,RCVext)
extRCV <- terra::extract(sentRCV, ptsRCV)[,-1] # REMOVE ID
sampdataRCV <- data.frame(class = ptsRCV$class, extRCV)
```

### Training the CART model (10 m) and plot the tree
Again, we can plot the regression tree for the model using 10 m data (Figure \@ref(fig:imgCART10mTree)).
```{r imgCART10mTree, fig.cap="10 m CART regression tree"}
library(rpart); library(rpart.plot)
cartmodel <- rpart(as.factor(class)~., data = sampdataRCV, 
                   method = 'class', minsplit = 5)
rpart.plot(cartmodel, fallen.leaves=F)
```

### Prediction using the CART model (10 m)
\index{prediction using the 10 m Sentinel-2 CART model}As with the 20 m prediction, we'll start by producing a composite of all class probabilities by pixel (Figure \@ref(fig:imgCART10mProb))...
```{r imgCART10mProb, fig.cap=paste0("CART classification, probabilities of each class, Sentinel-2 10 m ",imgDate)}
CARTpred <- predict(sentRCV, cartmodel, na.rm = TRUE)
plot(CARTpred)
```

```{r}
LC10m <- which.max(CARTpred)
cls <- names(CARTpred)
df <- data.frame(id = 1:length(cls), class=cls)
levels(LC10m) <- df
```

... and then map the class with the highest probability for each pixel (Figure \@ref(fig:imgCART10mHighestProb)).

```{r eval=F, include=F}
plot(LC10m, col=LCcolors)
```

```{r imgCART10mHighestProb, fig.height=6.5, fig.width=6, fig.cap=paste0("CART classification, highest probability class, Sentinel-2 10 m ",imgDate)}
  tm_shape(LC10m) + 
    tm_raster(col="class",  
              col.scale=tm_scale_categorical(values=LCcolors),
              col.legend=tm_legend(frame=F)) +
    tm_graticules(alpha=0.5, col="gray")
```


#### Cross-validation : 10 m Sentinel-2 CART model
\index{cross validation 10 m Sentinel-2}
```{r}
set.seed(42)
k <- 5 # number of folds
j <- sample(rep(1:k, each = round(nrow(sampdataRCV))/k))
x <- list()
for (k in 1:5) {
    train <- sampdataRCV[j!= k, ]
    test <- sampdataRCV[j == k, ]
    cart <- rpart(as.factor(class)~., data=train, method = 'class',
                  minsplit = 5)
    pclass <- predict(cart, test, na.rm = TRUE)
    # assign class to maximum probablity
    pclass <- apply(pclass, 1, which.max)
    # create a data.frame using the reference and prediction
    x[[k]] <- cbind(test$class, as.integer(pclass))
}
```

#### Accuracy metrics : 10 m Sentinel-2 CART model
As with the 20 m model, we'll look at the confusion matrix and the various accuracy metrics.

**Confusion Matrix (10 m)**
\index{confusion matrix Sentinel 10 m}
```{r}
y <- do.call(rbind, x)
y <- data.frame(y)
colnames(y) <- c('observed', 'predicted')
conmat <- table(y) # confusion matrix
```
```{r}
# change the name of the columns and rows to LCclass
if(length(LCclass)==dim(conmat)[2]){
  colnames(conmat) <- LCclass}else{
    print("Confusion matrix column dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
if(length(LCclass)==dim(conmat)[1]){
  rownames(conmat) <- LCclass}else{
    print("Confusion matrix row dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
```
```{r}
print(conmat)
```

**Overall Accuracy (10 m)**
```{r}
n <- sum(conmat) # number of total cases
diag <- diag(conmat) # number of correctly classified cases per class
OA10 <- sum(diag) / n # overall accuracy
OA10
```

**kappa (10 m)**

```{r}
rowsums <- apply(conmat, 1, sum) 
p <- rowsums / n # observed (true) cases per class
colsums <- apply(conmat, 2, sum)
q <- colsums / n # predicted cases per class
expAccuracy <- sum(p*q)
kappa10 <- (OA10 - expAccuracy) / (1 - expAccuracy)
kappa10
```

**User and Producer accuracy (10 m)**
```{r}
PA <- diag / colsums # Producer accuracy
UA <- diag / rowsums # User accuracy
outAcc <- data.frame(producerAccuracy = PA, userAccuracy = UA)

if(length(PA)==length(UA)){
  outAcc <- data.frame(producerAccuracy = PA, userAccuracy = UA)
  print(outAcc)}else{
     print("length(PA)!=length(UA)")
     print(PA)
     print(UA)
  }
```


## Classification Using Multiple Images Capturing Phenology

\index{phenological change}Most plants change their spectral response seasonally, a process referred to as *phenology*. In these montane meadows, phenological changes start with snow melt, followed by "green-up", then senescence. 2019 was a particularly dry year, so the period from green-up to senescence was short. We'll use a 3 June image as green-up, then a 28 June image as senescence.

**Spring**
```{r}
library(terra); library(stringr); library(igisci)
imgFolderSpring <- paste0("S2B_MSIL2A_20210603T184919_N0300_R113_T10TGK_20210603",
                     "T213609.SAFE\\GRANULE\\L2A_T10TGK_A022161_20210603T185928")
img10mFolderSpring <- paste0("~/sentinel2/",imgFolderSpring,"\\IMG_DATA\\R10m")
imgDateTimeSpring <- str_sub(imgFolderSpring,12,26)
```
**Summer**
```{r}
imgFolderSummer <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628",
                      "T230915.SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img10mFolderSummer <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R10m")
imgDateTimeSummer <- str_sub(imgFolderSummer,12,26)
```

```{r}
sentinelBands <- paste0("B",c("02","03","04","08"))
sentFilesSpring <- paste0(img10mFolderSpring,"/T10TGK_",
                              imgDateTimeSpring,"_",sentinelBands,"_10m.jp2")
sentinelSpring <- rast(sentFilesSpring)
names(sentinelSpring) <- paste0("spring",sentinelBands)

sentFilesSummer <- paste0(img10mFolderSummer,"/T10TGK_",
                              imgDateTimeSummer,"_",sentinelBands,"_10m.jp2")
sentinelSummer <- rast(sentFilesSummer)
names(sentinelSummer) <- paste0("summer",sentinelBands)
```

```{r}
RCVext <- ext(715680,725040,4419120,4429980)
sentRCVspring <- crop(sentinelSpring,RCVext)
sentRCVsummer <- crop(sentinelSummer,RCVext)
```

```{r}
blueSp <- sentRCVspring$springB02
greenSp <- sentRCVspring$springB03
redSp <- sentRCVspring$springB04
nirSp <- sentRCVspring$springB08
ndgiSp <- (0.616 * greenSp + 0.384 * nirSp - redSp)/
          (0.616 * greenSp + 0.384 * nirSp + redSp)
blueSu <- sentRCVsummer$summerB02
greenSu <- sentRCVsummer$summerB03
redSu <- sentRCVsummer$summerB04
nirSu <- sentRCVsummer$summerB08
ndgiSu <- (0.616 * greenSu + 0.384 * nirSu - redSu)/
          (0.616 * greenSu + 0.384 * nirSu + redSu)
```

### Create a 10-band stack from both images
Since we have two imagery dates and also a vegetation index (NDGI) for each, we have a total of 10 variables to work with. So we'll put these together as a multi-band stack simply using `c()` with the various bands. 
```{r}
sentRCV <- c(blueSp,greenSp,redSp,nirSp,ndgiSp,blueSu,greenSu,redSu,nirSu,ndgiSu)
names(sentRCV) <- c("blueSp","greenSp","redSp","nirSp","ndgiSp",
                    "blueSu","greenSu","redSu","nirSu","ndgiSu")
```

### Extract the training data (10 m spring + summer)
We'll repeat some of the steps where we read in the training data and set up the land cover classes, reducing the 7 to 6 classes, then extract from our new 10-band image stack the training data. 
```{r}
library(igisci)
trainPolys <- vect(ex("RCVimagery/train_polys7.shp"))
if (doCentroids) {ptsRCV <- centroids(trainPolys)
            }else{ptsRCV <- spatSample(trainPolys, 1000, method="regular")}
LCclass <- c("forest","hydric","mesic","rocky","water","willow","xeric")
classdf <- data.frame(value=1:length(LCclass),names=LCclass)
trainPolys <- vect(ex("RCVimagery/train_polys7.shp"))
trainPolys[trainPolys$class=="willow"]$class <- "hydric"
ptsRCV <- spatSample(trainPolys, 1000, method="random")
LCclass <- c("forest","hydric","mesic","rocky","water","xeric")
classdf <- data.frame(value=1:length(LCclass),names=LCclass)

extRCV <- terra::extract(sentRCV, ptsRCV)[,-1] # REMOVE ID
sampdataRCV <- data.frame(class = ptsRCV$class, extRCV)
```

### CART model and prediction (10 m spring + summer)
\index{imagery classification from phenological change CART model 10 m spring + summer}
```{r}
library(rpart)
# Train the model
cartmodel <- rpart(as.factor(class)~., 
                   data = sampdataRCV, method = 'class', minsplit = 5)
```

The CART model tree for the two-season 10 m example (Figure \@ref(fig:imgCARTtreePheno)) interestingly does not end up with anything in the "rocky" pixels...  

```{r imgCARTtreePheno, fig.cap="CART decision tree, Sentinel 10-m, spring and summer 2021 images"}
library(rpart.plot)
rpart.plot(cartmodel, fallen.leaves=F)
```

... and we can see in the probability map set that all "rocky" pixel probabilities are very low, below 0.03 (Figure \@ref(fig:imgCARTprobmapPheno)) ...  

```{r imgCARTprobmapPheno, fig.cap="CART classification, probabilities of each class, Sentinel-2 10 m, 2021 spring and summer phenology"}
CARTpred <- predict(sentRCV, cartmodel, na.rm = TRUE)
plot(CARTpred)
```

```{r}
LCpheno <- which.max(CARTpred)
cls <- names(CARTpred)
df <- data.frame(id = 1:length(cls), class=cls)
levels(LCpheno) <- df
```

... and thus no "rocky" classes are assigned (Figure \@ref(fig:imgCARThighestPheno)).

```{r eval=F}
LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
plot(LCpheno, col=LCcolors)
```

```{r imgCARThighestPheno, fig.height=6.5, fig.width=6, fig.cap="CART classification, highest probability class, Sentinel-2 10 m, 2021 spring and summer phenology"}
LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
tm_shape(LCpheno) + 
    tm_raster(col="class",  
            col.scale=tm_scale_categorical(values=LCcolors),
            col.legend=tm_legend(frame=F)) +
  tm_graticules(alpha=0.5, col="gray")

```



We might compare this plot with the two previous -- the 20 m (Figure \@ref(fig:imgCARThighestProb20m)) and 10 m (Figure \@ref(fig:imgCART10mHighestProb)) classifications from 28 June.


#### Cross validation : CART model for 10 m spring + summer
```{r}
set.seed(42)
k <- 5 # number of folds
j <- sample(rep(1:k, each = round(nrow(sampdataRCV))/k))
# table(j)
```

```{r}
x <- list()
for (k in 1:5) {
    train <- sampdataRCV[j!= k, ]
    test <- sampdataRCV[j == k, ]
    cart <- rpart(as.factor(class)~., data=train, method = 'class',
                  minsplit = 5)
    pclass <- predict(cart, test, na.rm = TRUE)
    # assign class to maximum probablity
    pclass <- apply(pclass, 1, which.max)
    # create a data.frame using the reference and prediction
    x[[k]] <- cbind(test$class, as.integer(pclass))
}
```

#### Accuracy metrics : CART model for 10 m spring + summer
Finally, as we did above, we'll look at the various accuracy metrics for the classification using the two-image (peak-growth spring and senescent summer) source.

**Confusion Matrix**
```{r}
y <- do.call(rbind, x)
y <- data.frame(y)
colnames(y) <- c('observed', 'predicted')
conmat <- table(y) # confusion matrix
```
```{r}
# change the name of the columns and rows to LCclass
if(length(LCclass)==dim(conmat)[2]){
  colnames(conmat) <- LCclass}else{
    print("Confusion matrix column dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
if(length(LCclass)==dim(conmat)[1]){
  rownames(conmat) <- LCclass}else{
    print("Confusion matrix row dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
```
```{r}
print(conmat)
```

**Overall Accuracy**
```{r}
n <- sum(conmat) # number of total cases
diag <- diag(conmat) # number of correctly classified cases per class
OA10ph <- sum(diag) / n # overall accuracy
OA10ph
```

**kappa**

```{r}
rowsums <- apply(conmat, 1, sum) 
p <- rowsums / n # observed (true) cases per class
colsums <- apply(conmat, 2, sum)
q <- colsums / n # predicted cases per class
expAccuracy <- sum(p*q)
kappa10ph <- (OA10ph - expAccuracy) / (1 - expAccuracy)
kappa10ph
```




## Classification Using Multiple Images Capturing Phenology : 20m

To apply the concept of using imagery from two different times (we'll call spring and summer), we'll use 9 bands each two  Sentinel-2 images, at 20 m resolution. As with the 10 m Sentinel imagery, we'll combine an early and late June images from 2021, a particularly dry year with limited snow pack and rapid senescence.  In these montane meadows, phenological changes start with snow melt, followed by "green-up", then senescence. 

```{r}
library(stringr)
library(terra)
imgFolder <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628T230915.",
                    "SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img20mFolder <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R20m")
imgDateTime <- str_sub(imgFolder,12,26)
imgDateTime
imgDate <- str_sub(imgDateTime,1,8)
```

```{r}
sentinelBands <- paste0("B",c("02","03","04","05","06","07","8A","11","12"))
sentFiles <- paste0(img20mFolder,"/T10TGK_",imgDateTime,"_",
                        sentinelBands,"_20m.jp2")
sentinel <- rast(sentFiles)
sentinelBands
names(sentinel) <- sentinelBands
```

**Spring**
```{r}
library(terra); library(stringr); library(igisci)
imgFolderSpring <- paste0("S2B_MSIL2A_20210603T184919_N0300_R113_T10TGK_20210603",
                     "T213609.SAFE\\GRANULE\\L2A_T10TGK_A022161_20210603T185928")
img10mFolderSpring <- paste0("~/sentinel2/",imgFolderSpring,"\\IMG_DATA\\R20m")
imgDateTimeSpring <- str_sub(imgFolderSpring,12,26)
```
**Summer**
```{r}
imgFolderSummer <- paste0("S2A_MSIL2A_20210628T184921_N0300_R113_T10TGK_20210628",
                      "T230915.SAFE\\GRANULE\\L2A_T10TGK_A031427_20210628T185628")
img10mFolderSummer <- paste0("~/sentinel2/",imgFolder,"\\IMG_DATA\\R20m")
imgDateTimeSummer <- str_sub(imgFolderSummer,12,26)
```

Since we're going to combine two images, we'll need unique band names, so we'll add "sp" to the earlier image band names (to produce "spB01", etc.), and "su" to the later image band names, and get 18 unique names.
```{r}
sentinelBands <- paste0("B",c("02","03","04","05","06","07","8A","11","12"))
sentFilesSpring <- paste0(img10mFolderSpring,"/T10TGK_",
                              imgDateTimeSpring,"_",sentinelBands,"_20m.jp2")
sentinelSpring <- rast(sentFilesSpring)
names(sentinelSpring) <- paste0("sp",str_sub(sentinelBands,2,3))

sentFilesSummer <- paste0(img10mFolderSummer,"/T10TGK_",
                              imgDateTimeSummer,"_",sentinelBands,"_20m.jp2")
sentinelSummer <- rast(sentFilesSummer)
names(sentinelSummer) <- paste0("su",str_sub(sentinelBands,2,3))
```

```{r}
RCVext <- ext(715680,725040,4419120,4429980)
sentRCVspring <- crop(sentinelSpring,RCVext)
sentRCVsummer <- crop(sentinelSummer,RCVext)
```


### Create a 18-band stack from both images
Since we have two imagery dates , we have a total of 18 variables to work with. So we'll put these together as a multi-band stack simply using `c()` with the two 9-band images to get an 18-band stack. 
```{r}
sentRCV <- c(sentRCVspring, sentRCVsummer)
```

```{r}
plotRGB(subset(sentRCV,c(7,3,2)),stretch="lin")
```

The "summer" bands are 10 through 18, so the following image display of NIR-R-G specified by numbers c(16,12,11) is comparable to c(7,3,2) from spring.
```{r}
plotRGB(subset(sentRCV,c(16,12,11)),stretch="lin")
```
We haven't looked at the spectral signature for the spring image, so we'll look at a graph of all 18 bands from the combination, using the 7-class training set that includes willows as separate from hydric.

```{r imgSpectralSignatures7pheno, fig.cap=paste0("Spectral signature of 7-level training polygons, 20 m Sentinel-2 imagery from ",imgDate)}
trainPolys7 <- vect(ex("RCVimagery/train_polys7.shp"))
if (doCentroids) {ptsRCV <- centroids(trainPolys7)
            }else{ptsRCV <- spatSample(trainPolys7, 1000, method="regular")}
extRCV <- terra::extract(sentRCV, ptsRCV)
specRCV <- aggregate(extRCV[,-1], list(ptsRCV$class), mean)
specRCVdf <- specRCV|> rename(cover=Group.1)
coverColors <- c("darkgreen","cyan", "orange","gray","blue","green","red")
pivot_specRCVdf <- specRCVdf |>
  pivot_longer(cols=sp02:su12, names_to="band", values_to="reflectance")
ggplot(data=pivot_specRCVdf, aes(x=band,y=reflectance,col=cover,group=cover)) + geom_line(size=1.5) +
  labs(y="Reflectance 12 bit") +
  scale_color_discrete(type=coverColors)
  
```



### Extract the training data (20 m spring + summer)
We'll repeat some of the steps where we read in the training data and set up the land cover classes, leaving the 7 classes, then extract from our new 18-band image stack. 
```{r}
library(igisci)
trainPolys <- vect(ex("RCVimagery/train_polys7.shp"))
if (doCentroids) {ptsRCV <- centroids(trainPolys)
            }else{ptsRCV <- spatSample(trainPolys, 1000, method="regular")}
LCclass <- c("forest","hydric","mesic","rocky","water","willow","xeric")
classdf <- data.frame(value=1:length(LCclass),names=LCclass)
extRCV <- terra::extract(sentRCV, ptsRCV)[,-1] # REMOVE ID
sampdataRCV <- data.frame(class = ptsRCV$class, extRCV)
```

### CART model and prediction (20 m spring + summer)
\index{imagery classification from phenological change CART model 20 m spring + summer}
```{r}
library(rpart)
# Train the model
cartmodel <- rpart(as.factor(class)~., 
                   data = sampdataRCV, method = 'class', minsplit = 5)
```

The CART model tree for the two-season 20 m example (Figure \@ref(fig:imgCARTtreePheno20)), in contrast with the 10 m classification, ends up representing "rocky" with classified pixels (Figures \@ref(fig:imgCARTprobmapPheno20) and \@ref(fig:imgCARThighestPheno20)).  

```{r imgCARTtreePheno20, fig.cap="CART decision tree, Sentinel 20-m, spring and summer 2021 images"}
library(rpart.plot)
rpart.plot(cartmodel, fallen.leaves=F, box.palette=list("Greens","Purples","Oranges","Grays","Blues","Browns","Reds"))
```

```{r imgCARTprobmapPheno20, fig.cap="CART classification, probabilities of each class, Sentinel-2 20 m, 2021 spring and summer phenology"}
CARTpred <- predict(sentRCV, cartmodel, na.rm = TRUE)
plot(CARTpred)
```

```{r}
LCpheno <- which.max(CARTpred)
cls <- names(CARTpred)
df <- data.frame(id = 1:length(cls), class=cls)
levels(LCpheno) <- df
```

```{r eval=F}
# LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
LCcolors <- c("darkgreen","cyan", "orange","gray","blue","green","red")
plot(LCpheno, col=LCcolors)
```

```{r imgCARThighestPheno20, fig.height=6.5, fig.width=6, fig.cap="CART classification, highest probability class, Sentinel-2 20 m, 2021 spring and summer phenology"}
library(tmap)
# LCcolors <- c("green4","cyan","gold","grey90","mediumblue","red")
LCcolors <- c("green4","cyan", "gold","gray90","mediumblue","green","red")
tm_shape(LCpheno) + 
    tm_raster(col="class",  
            col.scale=tm_scale_categorical(values=LCcolors),
            col.legend=tm_legend(frame=F)) +
  tm_graticules(alpha=0.5, col="gray")

```



We might compare this plot with the previous 20 m (Figure \@ref(fig:imgCARThighestProb20m)), 10 m (Figure \@ref(fig:imgCART10mHighestProb)), and 10 m spring+summer classification (\@ref(fig:imgCARThighestPheno)).


#### Cross validation : CART model for 20 m spring + summer
```{r}
set.seed(42)
k <- 5 # number of folds
j <- sample(rep(1:k, each = round(nrow(sampdataRCV))/k))
# table(j)
```

```{r}
x <- list()
for (k in 1:5) {
    train <- sampdataRCV[j!= k, ]
    test <- sampdataRCV[j == k, ]
    cart <- rpart(as.factor(class)~., data=train, method = 'class',
                  minsplit = 5)
    pclass <- predict(cart, test, na.rm = TRUE)
    # assign class to maximum probablity
    pclass <- apply(pclass, 1, which.max)
    # create a data.frame using the reference and prediction
    x[[k]] <- cbind(test$class, as.integer(pclass))
}
```

#### Accuracy metrics : CART model for 20 m spring + summer
Finally, as we did above, we'll look at the various accuracy metrics for the classification using the two-image (peak-growth spring and senescent summer) source.

**Confusion Matrix**
```{r}
y <- do.call(rbind, x)
y <- data.frame(y)
colnames(y) <- c('observed', 'predicted')
conmat <- table(y) # confusion matrix
```
```{r}
# change the name of the columns and rows to LCclass
if(length(LCclass)==dim(conmat)[2]){
  colnames(conmat) <- LCclass}else{
    print("Confusion matrix column dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
if(length(LCclass)==dim(conmat)[1]){
  rownames(conmat) <- LCclass}else{
    print("Confusion matrix row dimension/LCclass array mismatch")
    print(str(LCclass))
    print(str(conmat))
  }
```
```{r}
print(conmat)
```

**Overall Accuracy**
```{r}
n <- sum(conmat) # number of total cases
diag <- diag(conmat) # number of correctly classified cases per class
OA20ph <- sum(diag) / n # overall accuracy
OA20ph
```

**kappa**

```{r}
rowsums <- apply(conmat, 1, sum) 
p <- rowsums / n # observed (true) cases per class
colsums <- apply(conmat, 2, sum)
q <- colsums / n # predicted cases per class
expAccuracy <- sum(p*q)
kappa20ph <- (OA20ph - expAccuracy) / (1 - expAccuracy)
kappa20ph
```

**User and Producer accuracy**
```{r}
PA <- diag / colsums # Producer accuracy
UA <- diag / rowsums # User accuracy
if(length(PA)==length(UA)){
  outAcc <- data.frame(producerAccuracy = PA, userAccuracy = UA)
  print(outAcc)}else{
     print("length(PA)!=length(UA)")
     print(PA)
     print(UA)
  }

```




```{r}
stats <- tribble(~classification, ~OA, ~Kappa,
                 "20m", round(OA20,3), round(kappa20,3),
                 "10m", round(OA10,3), round(kappa10,3),
                 "10m_2season", round(OA10ph,3), round(kappa10ph,3),
                 "20m_2season", round(OA20ph,3), round(kappa20ph,3),
)
DT::datatable(stats)

```


The overall accuracy (OA) and kappa statistic suggest high accuracy in all of these classifications, given these training data (which is mostly captured in meadow locations, not the adjacent hillslopes other than clearly forested areas apparent in all imagery), though none are perfect. My impression from knowing the area, however, points to both of the 20 m classifications as having the best representation, and the two-season 18-band result looks the best to my eye.

## Conclusions and Next Steps for Imagery Classification

\index{imagery classification next steps}In this chapter, we've seen that we can effectively classify imagery (at least produce convincing maps!), but you can probably sense from the varying results that classification isn't always as consistent as we might expect. We saw something similar with interpolation models; you can get very different results with different methods, and so you need to consider what method and settings might be most appropriate, based on the nature of your data -- its scale and density, for instance -- and the goals of your research. 

The next steps might be to compare some of these classifications with our field data (basically go back to the training set), but I would be tempted to do that in ArcGIS, ERDAS, or maybe QGIS after writing out our classification rasters to geotiffs. We could then also take advantage of the better cartographic capabilities of GIS and remote sensing software when working with multi-band imagery. 

We've really only scratched the surface on *remote sensing* methods for imagery classification, and as noted above, readers are encouraged to enroll in remote sensing classes at your local university to learn more of its theory and relevant technology. For instance, why did we look at Sentinel-2, and not Landsat, PlanetScope, ASTER, Spot, ... (the list goes on)? There are lots of satellites out there, more all the time, and understanding the the spectral-sensing capabilities of their sensors, and the spatial and temporal properties of their missions, is essential for making the right choice for your research. 

There's also the consideration of relative cost, both for the imagery itself and for your time in processing it. We chose Sentinel-2 because it's free and served the purpose well, since the 10 and 20 m resolution worked at the scale of the study area in question. It also actually classifies more readily than finer resolution imagery. But there's a lot more to that decision, and we also employ 1 and 5 cm pixel imagery from drone surveys of the same study area, but for a different purpose (erosion studies) and at much lower image capture frequency, given the significant field time required.

Finally, we focused on pixel classifiers. One advantage of these is that they're reasonably similar to general classification models, whether using machine learning or not. But there are other imagery classification methods developed in the remote-sensing as well as medical-imaging communities, such as object-based classifiers which use vector geometry to aid in the classification of features. As with so much else in remote sensing methods, we'll leave those and other methods for you to explore.

\pagebreak
## Exercises: Imagery Analysis and Classification Models

The exercises for this chapter are to apply some of the same methods above to your own downloaded Sentinel2 scene. We won't use training sets for spectral signatures or supervised classification, but that would be the next step once you've collected training data, either in the field, from high-resolution imagery, or from a combination of sources.

:::{.exercise}
Download another Sentinel-2 image from Copernicus (you'll need to create a free account to get SciHub credentials), and use it to create RGB and NIR-R-G image displays of the entire imagery scene. 

For downloading (and other processing), you might also try the `sen2r` package. You'll also need SciHub credentials, which you can get from https://scihub.copernicus.eu/. And while `sen2r` will download the entire images without it, Google Cloud SDK must be installed (see https://cloud.google.com/sdk/docs/install), which in turn will require a Python installation to do other processing.

**Caution**: These files can be very large, with each scene about a gigabyte, so be careful to limit how many images you're downloading. It's a good idea to be selective about which scenes to acquire, and the Copernicus Open Access Hub is a good interface for using visual inspection, which helps you decide if an image shows your area of interest well. For instance, while automated selection of percentage cloud cover is useful as a first step, your area of interest may be an anomaly: even a 50% cloud coverage for the overall scene may not cover your area of interest, or a 5% cloud coverage may sit right over it.
:::

:::{.exercise}
From your imagery, find coordinates to crop with for an area of interest, then as we've done for Red Clover Valley, create clipped RGB and NIR-R-G displays.
:::

:::{.exercise}
Create and display NDVI and NDGI indices from your cropped image. 
:::

:::{.exercise}
From the same cropped image, create a k-means unsupervised classification, with your choice of number of classes. 
:::
